Type-checking Linearity in GHC Core

### 2024-01-23 icfp.tex
stopped at newMArray

\documentclass[acmsmall, screen, review]{acmart}
\PassOptionsToPackage{dvipsnames}{xcolor}
\usepackage{boldline}
\usepackage{xcolor}
\usepackage{xargs}
\usepackage{framed}
\usepackage[framemethod=default]{mdframed}
\usepackage{mathpartir}
\usepackage{mathtools}
\usepackage{soul}
\usepackage{cmll} % ROMES:TODO: not allowed?
\usepackage{amsthm}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{cleveref}
\usepackage{makecell}
\usepackage{fancyvrb}
\usepackage{thmtools, thm-restate}
\usepackage{listings}

@@

@@

\lstset{
  frame=none,
  xleftmargin=2pt,
  stepnumber=1,
  numbers=left,
  numbersep=5pt,
  numberstyle=\ttfamily\tiny\color[gray]{0.3},
  belowcaptionskip=\bigskipamount,
  captionpos=b,
  escapeinside={*'}{'*},
  language=haskell,
  tabsize=2,
  emphstyle={\bf},
  commentstyle=\it,
  stringstyle=\mdseries\rmfamily,
  showspaces=false,
  keywordstyle=\bfseries\rmfamily,
  columns=flexible,
  basicstyle=\small\sffamily,
  showstringspaces=false,
  morecomment=[l]\%,
}

@@

Despite their promise and extensive presence in research
literature~\cite{Wadler1990LinearTC,CERVESATO2000133,10.1093/logcom/2.3.297},
an effective design combining linear and non-linear typing is both
challenging and necessary to bring the advantages of linear typing to
mainstream languages.
%
Consequently, few general purpose programming languages have linear
type systems. Among them are Idris~2~\cite{brady:LIPIcs.ECOOP.2021.9},
% a linearly and dependently typed language based on Quantitative Type Theory,
Rust~\cite{10.1145/2692956.2663188}, a language whose
ownership types build on linear types to guarantee memory safety
without garbage collection or reference counting, and, more recently,
Haskell~\cite{cite:linearhaskell}, a pure, functional, and
\emph{lazy} general purpose programming language.
%
% Besides Haskell's supporting linear
% types according to the said paper, Idris 2\cite{} supports linear types in a
% dependently typed setting, Clean\cite{} has uniqueness types which are closely
% related to linear types, and Rust\cite{} has ownership types which build from
% linear types. 
%
Linearity in Haskell stands out from linearity in other languages for the
following reasons, essential to our work:

\begin{itemize}
    % \item Linear types were only added to the language roughly \emph{31 years
    %     after} Haskell's inception, unlike Rust and Idris 2 which were designed
    %     with linearity from the start. It is an especially difficult endeavour
    %     to add linear types to a well-established language with a large library
    %     ecosystem and many active users, rather than to develop the language
    %     from the ground up with linear types in mind. The successful
    %     approach as implemented in the Glasgow Haskell Compiler (GHC), the leading
    %     Haskell compiler, is based on Linear Haskell~\cite{cite:linearhaskell}, which describes a linear type system designed
    %     with retaining backwards-compatibility and allowing code reuse across
    %     linear and non-linear users of the same libraries.
    %     We describe Linear Haskell in detail in Section~\ref{sec:linear-haskell}.

    \item Linear types permeate Haskell down to (its) Core, the
        intermediate language into which Haskell is
        translated. Core is a minimal, explicitly typed,
        functional language, to which multiple
        Core-to-Core optimising transformations are
        applied during compilation. Due to Core's minimal design, typechecking
        Core programs is very efficient and doing so serves as a sanity check to the
        correction of the source transformations. If the resulting optimised
        Core program fails to typecheck, the optimising
        transformations are very likely to have introduced an error
        in the resulting program. We present Core and its formal
        basis, System~$F_C$~\cite{cite:systemfc}, in Section~\ref{sec:core}.
        % TODO: \item values in rust are linear by default while non-linear is
        % the haskell default?

    \item Both Haskell and its intermediate language Core are \emph{lazily}
        evaluated, i.e., expressions in Haskell are only evaluated when needed,
        unlike C or Rust, which are \emph{eagerly} evaluated.
        %
        Laziness allows an optimising compiler to aggressively transform the source
        program without changing its semantics and, indeed, the Glasgow Haskell
        Compiler (GHC) heavily transforms Core by leveraging its laziness.
        %
        However, lazy evaluation interacts non-trivially with linearity.
        Intuitively, since expressions are not necessarily evaluated, an
        occurrence of a linear resource in an expression does not necessarily
        entail consuming that resource (i.e., if the expression is not evaluated,
        the resource is not used).

        In eagerly evaluated languages, the distinction between syntactic uses of a
        resource and the actual use of linear resources at runtime does not exist --
        an occurrence of a variable in the program always entails consuming it.
        %
        This interaction is unique to Haskell since, as far as we know, it is the
        only language featuring both laziness and linearity.
        %
        We review lazy and eager evaluation strategies in Section~\ref{sec:background:evaluation-strategies}.

\end{itemize}
%

@@

In spite of a linearly typed Core ultimately being the desired intermediate
language for the Glasgow Haskell Compiler,
%ever since Linear Haskell was implemented,
both in its expressiveness to completely represent a Haskell with
linear types and in enabling more extensive compiler validations plus improved optimisations,
% it is currently impossible to enable linearity checking in Core. The reason is
linearity is effectively ignored in Core. The reason is not evidently clear:
if we can typecheck linearity in the surface level Haskell, it should also be
possible, and natural, to do so in Core.
% why do we fail to do so in Core?

@@

\end{itemize}
%
We review background concepts fundamental to our work in
Chapter~\ref{sec:background}, including linear type systems, linear types in
Haskell, evaluation strategies, Core's type system, and multiple optimising
transformations applied by GHC in its compilation process.
%
We compare our contributions to related work and discuss possible
avenues for further research (highlighting so-called \emph{multiplicity
coercions}) in Chapter~\ref{sec:discussion}, which concludes the document.

@@

Consider the following program in a functional Haskell-like language,
where a computation that closes the given |handle| is bound to |x| before the
|handle| is returned:
%
\begin{code}
f : Handle \lolli Handle
f handle = let x = close handle in handle
\end{code}
%
In this seemingly innocent example, the |handle| appears to be closed before
returned, whereas in fact the handle will only be closed if the let bound
computation is effectively run (i.e. evaluated).
%
The example illustrates that \emph{consuming} a resource is not necessarily
synonymous with using it syntactically, as depending on the evaluation strategy
of the language, the computation that closes the handle might or not be
evaluated, and if it isn't, the |handle| in that unused computation
is not consumed.
%
Expanding on this, consider the above example program under distinct
evaluation strategies:
%
\begin{description}

\item[Call-by-value] With \emph{eager} evaluation semantics, the let bound expression |close
handle| is eagerly evaluated, and the |handle| will be closed before being
returned. It is clear that a linear type system should not accept such a
program since the linear resource |handle| is duplicated -- it is used in a
computation that closes it, while still being made available to the caller of
the function.

\item[Call-by-need] On the other hand, with \emph{lazy} evaluation semantics, the let
bound expression will only be evaluated when the binding |x| is needed. We return
the |handle| right away, and the let binding is forgotten as it cannot be
used outside the scope of the function, so the handle is not closed by |f|.
Under the lens of \emph{call-by-need} evaluation, \emph{using} a resource in a
let binding only results in the resource being \emph{consumed} if the binding
itself is \emph{consumed}. We argue that a linear type system under
\emph{call-by-need} evaluation should accept the above program, unlike a linear
type system for the same program evaluated \emph{call-by-value}.

\end{description}
%
%Rather, \emph{consuming} a resource is intrinsically defined by the evaluation model of the language.
%
% Under distinct evaluation strategies, the above example (and programs in general) have different semantics:
%
%
Intuitively, a computation that depends on a linear resource to produce a
result consumes that resource iff the result is effectively computed; in
contrast, a computation that depends on a linear resource, but is never run,
will not consume that resource.

From this observation, and exploring the connection between computation and evaluation,
% \todo{Alguma dificuldade em dizer exatamente como é que evaluation drives/is computation}
it becomes clear that \emph{linearity} and \emph{consuming resources}, in the
above example and for programs in general, should be defined in function of the
language's evaluation strategy.
%
We turn our focus to \emph{linearity} under \emph{call-by-need}, not only
because GHC Core is \emph{call-by-need}, but also because the distinction
between semantically and syntactically consuming a resource is only exposed
under \emph{non-strict} semantics.
%
Indeed, under \emph{call-by-value}, syntactic occurrences of a linear resource
directly correspond to semantically using that resource\footnote{With the minor exception
of trivial aliases, which don't entail any computation even in
\emph{call-by-value}. In theory, we could use in mutual exclusion any of the
aliases to refer to a resource without loss of linearity} because \emph{all}
expressions are eagerly evaluated -- if all computations are eagerly run, all
linear resources required by computations are \emph{eagerly consumed}.

@@

\subsubsection{Let bindings}
We start our discussion with non-strict (non-recursive) let bindings, i.e. let bindings whose
body is evaluated only when the binding is needed, rather than when declared.
In Core, a let binding entails the creation of a \emph{thunk} that suspends the
evaluation of the let body (for background, see
Section~\ref{sec:background:evaluation-strategies}). When the binding itself is evaluated, the \emph{thunk} is
\emph{forced} and the evaluation is carried out. The result overwrites the
\emph{thunk} -- the let binding now points to the result of the evaluation.
% A \emph{thunk} is \emph{forced} (and the suspended computation is evaluated) when
% the binding itself is evaluated.

In a linear type system, a non-strict let binding that depends on a linear
resource |x| doesn't consume the resource as long as the binding isn't
evaluated -- the suspended computation only uses the resource if it is run. For
this reason, we can't naively tell whether |x| is consumed just by looking at
the let binding body. In the following example, we assign a computation that
depends on the resource |x| to a binder, which is then returned:
%
\begin{notyet}
\begin{code}
f1 :: (a -o b) -> a -o b
f1 use x =
  let y = use x
   in y
\end{code}
\end{notyet}
%
The linear resource |x| is used exactly once, since it is used exactly once in
the body of the binding and the binding is used exactly once in the let body.
%
According to Linear Haskell's core calculus~$\lambda_{\to}^{q}$~\cite{cite:linearhaskell}, let
bound variables are annotated with a multiplicity which is multiplied (as per
the multiplicity semiring) by the multiplicity of all variables that are free
in the binder's body.
%
In short, if a let binder is linear (has
multiplicity $1$) then the linear variables free in its body are only used
once; if the let binder is unrestricted (has multiplicity $\omega$) then the
resources in its body are consumed many times, meaning no linear variables can
occur in that let binder's body.
%
Unfortunately, GHC's implementation of Linear
Haskell doesn't seem to infer multiplicities for lets yet, so while the above
program should typecheck in Linear Haskell, it is rejected by GHC.

The next example exposes the case in which the let binder is ignored in the let
body. Here, the linear resource |x| is used in |y|'s body and in the let
body, however, the resource is still used semantically linearly because |y| isn't used at all, thus |x| is consumed just once in the let body:
%
\begin{notyet}
\begin{code}
f2 :: (a -o a) -> a -o a
f2 use x =
  let y = use x
  in use x
\end{code}
\end{notyet}
%
Programmers don't often write bindings that are completely unused, yet, an
optimising compiler will produce intermediate programs with unused bindings\footnote{Unused bindings are then
also dropped by the optimising compiler} 
from transformations such as inlining, which can substitute out occurrences
of the binder (e.g. |y| is inlined in the let body).

Let bindings can also go unused if they are defined before branching on case
alternatives. At runtime, depending on the branch taken, the let binding will
be evaluated only if it occurs in that branch.
Both optimising transformations (float-out), and programmers used to non-strict evaluation, can produce
programs with bindings that are selectively used in the case alternatives, for
instance:
%
\begin{notyet}
\begin{code}
f3 :: (a -o a) -> Bool -> a -o a
f3 use bool x =
  let y = use x
  in case bool of
      True -> x
      False -> y
\end{code}
\end{notyet}
%
This example essentially merges |f1| with |f2|, using |x| directly in one
branch and using |y| in the other. Semantically, this program is linear because
the linear resource |x| ends up being used exactly once in both case
alternatives, directly or indirectly.

Shifting our focus from not using a let binding to using it (more than once),
we reiterate that a let binding creates a \emph{thunk} which is only evaluated
once, and re-used subsequently. Despite the binder body only being evaluated
once, and thus its resources only used once to compute a result, we can still
only consume said result of the computation once -- perhaps surprisingly, as the
perception so far is that ``resources are consumed during computation'' and
multiple uses of the same let binder share the result that was computed only
once. Illustratively, the following program must \emph{not} typecheck:
%
\begin{noway}
\begin{code}
f4 :: (a -o b) -> a -o (b, b)
f4 use x =
  let y = use x
  in (y, y)
\end{code}
\end{noway}
%
Intuitively, the result of the computation must also be used exactly once,
despite being effectively computed just once, because said result may still
contain (parts of) the linear resource. The trivial example is |f4| applied to
|id| -- the result of computing |id x| is |x|, and |x| must definitely not be
shared! Indeed, if the result of the computation involving the linear resource
was, e.g., an unrestricted integer, then sharing the result would not involve consuming the
resource more than once.
% ROMES:TODO:!!!!
%\todo{this kind of hints that maybe somehow in the
%alternative scrutinizing a "trivial" atom we could make the let binding
%unrestricted after we were sure it was evaluated once, but that isn't easy
%without erasing too many things before, we kind of tried this once}
%
Concretely, the result of evaluating a let binder body using linear resources, if computed, must be
consumed exactly once, or, otherwise, we risk discarding or duplicating said resources.

Lastly, consider a program which defines two let bindings |z| and |y|, where
|z| uses |y| which in turn uses the linear resource |x|:
%
\begin{noway}
\begin{code}
f5 :: (a -o a) -> a -o ()
f5 use x =
  let y = use x
  let z = use y
  in ()
\end{code}
\end{noway}
%
Even though the binding |y| is used in |z|, |x| is still never consumed because
|z| isn't evaluated in the let body, and consequently |y| isn't evaluated
either -- never consuming |x|. We use this example to highlight that even for
let bound variables, the syntactic occurrence of a variable isn't enough to
determine whether it is used. Instead, we ought to think of uses of |y| as
implying using |x|, and therefore uses of |z| imply using |x|, however, if
neither is used, then |x| isn't used. Since |x| is effectively discarded, this
example also violates linearity.

% \parawith{Summary}
The examples so far build an intuition for semantic linearity in the presence
of lazy let bindings. In essence, an unused let binding doesn't consume any
resources, and a let binding used exactly once consumes its resources exactly
once. Let binders that depend on linear resources must be used \emph{at most
once} -- let bound variables are \emph{affine} in the let body.
%
Moreover, if the let binding ($y$) isn't used in the let body, then the
resources it depends on ($\ov{x}$) must still be used -- the binding $y$ is
mutually exclusive with the resources $\ov{x}$ (for the resources to be used
linearly, either the binder occurs exactly once $y$, or the resources $\ov{x}$
do). We'll later see how we can encode this principle of mutual exclusivity
between let bindings and their dependencies using so called \emph{usage
environments}, in Section~\ref{sec:usage-environments}.

\subsubsection{Recursive let bindings\label{sec:semantic-linearity-examples:recursive-lets}}
Second, we look into recursive let bindings. For the most part,
recursive let bindings behave as non-recursive let bindings, i.e. we must use them \emph{at
most once} because, when evaluated, the linear resources used in the binders
bodies are consumed. The defining property of a group of mutually recursive let
bindings is that the binders bound in that group can occur, without restrictions, in
the bodies of those same binders. The same way that, in a let body, evaluating
a binding that uses some resource exactly once consumes that resource once,
using the binding in its own definition also entails using that resource once.
Consider the following program, that calls a recursive let-bound function
defined in terms of the linear resource |x| and itself:
%
\begin{notyet}
\begin{code}
f6 :: Bool -> a -o a
f6 bool x =
  let go b
        = case b of
           True -> x
           False -> go (not b)
  in go bool
\end{code}
\end{notyet}
%
Function |f6| is semantically linear because, iff it is consumed exactly once,
then |x| is consumed exactly once. We can see this by case analysis on |go|'s argument
\begin{itemize}
\item When |bool| is |True|, we'll use the resource |x|
\item When |bool| is |False|, we recurse by calling |go| on |True|, which in turn will use the resource |x|.
\end{itemize}
In |go|'s body, |x| is used directly in one branch and indirectly in the
other, by recursively calling |go| (which we know will result in using |x|
linearly).
% ROMES:TODO:!!
%\todo{this bit is quite hard to explain. It is some sort of cyclic
%argument -- we kind of assume go uses x linearly s.t. when go itself is used
%then we're using x linearly. recursion...}
%
It so happens that |go| will terminate on any input, and will always consume
|x|. However, termination is not a requirement for a binding to use |x| linearly,
and we could have a similar example in which |go| might never terminate but still
uses |x| linearly if evaluated:
%
\begin{notyet}
\begin{code}
f7 :: Bool -> a -o a
f7 bool x =
  let go b
        = case b of
           True -> x
           False -> go b
  in go bool
\end{code}
\end{notyet}
%
The key to linearity in the presence of non-termination is Linear Haskell's
definition of a linear function: \emph{if a linear function application (|f u|) is
consumed exactly once, then the argument (|u|) is consumed exactly once}.
If |f u| doesn't terminate, it is never consumed, thus the claim holds
vacuously; that's why |f8| typechecks:
\begin{working}
\begin{code}
f8 :: a -o b
f8 x = f8 x
\end{code}
\end{working}
%
If |go| doesn't terminate, we aren't able to compute (nor consume) the result
of |f7|, so we do not promise anything about |x| being consumed (|f7|'s
linearity holds trivially). If it did terminate, it would consume |x| exactly
once (e.g. if |go| was applied to |True|).

>>>>

\documentclass[acmsmall, screen, review]{acmart}
\PassOptionsToPackage{dvipsnames}{xcolor}
\usepackage{boldline}
\usepackage{xcolor}
\usepackage{xargs}
\usepackage{framed}
\usepackage[framemethod=default]{mdframed}
\usepackage{mathpartir}
\usepackage{mathtools}
\usepackage{soul}
\usepackage{cmll} % ROMES:TODO: not allowed?
\usepackage{amsthm}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{cleveref}
\usepackage{makecell}
\usepackage{fancyvrb}
\usepackage{thmtools, thm-restate}
\usepackage{listings}
\usepackage{todonotes}
\usepackage[justification=centering]{subfig}

@@

\newcommand{\incode}[1]{\lstinline{#1}}

@@

\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.3,0}
\definecolor{darkpink}{rgb}{0.4,0,0.3}
\definecolor{graygreen}{rgb}{0.3,0.5,0.3}
\definecolor{grayblue}{rgb}{0.2,0.2,0.6}
\definecolor{grayred}{rgb}{0.5,0.2,0.2}

\lstset{
  frame=none,
  xleftmargin=2pt,
  stepnumber=1,
  %numbers=left,
  %numbersep=5pt,
  numberstyle=\ttfamily\tiny\color[gray]{0.3},
  belowcaptionskip=\bigskipamount,
  captionpos=b,
  escapeinside={*'}{'*},
  language=haskell,
  tabsize=2,
  emphstyle={\bf},
  commentstyle=\it,
  stringstyle=\mdseries\rmfamily,
  showspaces=false,
  keywordstyle=\bfseries\rmfamily\color{darkblue}, 
  columns=flexible,
  basicstyle=\small\sffamily,
  showstringspaces=false,
  morecomment=[l]\%,
}

@@

Despite their promise and extensive presence in research
literature~\cite{Wadler1990LinearTC,CERVESATO2000133,10.1093/logcom/2.3.297},
an effective design combining linear and non-linear typing is both
challenging and necessary to bring the advantages of linear typing to
mainstream languages.
%
Consequently, few general purpose programming languages have linear
type systems. Among them are Idris~2~\cite{brady:LIPIcs.ECOOP.2021.9},
% a linearly and dependently typed language based on Quantitative Type Theory,
Rust~\cite{10.1145/2692956.2663188}, a language whose
ownership types build on linear types to guarantee memory safety
without garbage collection or reference counting, and, more recently,
Haskell~\cite{cite:linearhaskell}, a pure, functional, and
\emph{lazy} general purpose programming language.
%
% Besides Haskell's supporting linear
% types according to the said paper, Idris 2\cite{} supports linear types in a
% dependently typed setting, Clean\cite{} has uniqueness types which are closely
% related to linear types, and Rust\cite{} has ownership types which build from
% linear types. 
%
Linearity in Haskell stands out from linearity in other languages:
%\begin{itemize}
    % \item Linear types were only added to the language roughly \emph{31 years
    %     after} Haskell's inception, unlike Rust and Idris 2 which were designed
    %     with linearity from the start. It is an especially difficult endeavour
    %     to add linear types to a well-established language with a large library
    %     ecosystem and many active users, rather than to develop the language
    %     from the ground up with linear types in mind. The successful
    %     approach as implemented in the Glasgow Haskell Compiler (GHC), the leading
    %     Haskell compiler, is based on Linear Haskell~\cite{cite:linearhaskell}, which describes a linear type system designed
    %     with retaining backwards-compatibility and allowing code reuse across
    %     linear and non-linear users of the same libraries.
    %     We describe Linear Haskell in detail in Section~\ref{sec:linear-haskell}.
%
%\item
  Linear types permeate Haskell down to (its) Core, the
        intermediate language into which Haskell is
        translated. Core is a minimal, explicitly typed,
        functional language, to which multiple
        Core-to-Core optimising transformations are
        applied during compilation. Due to Core's minimal design, typechecking
        Core programs is very efficient and doing so serves as a sanity check to the
        correction of the source transformations. If the resulting optimised
        Core program fails to typecheck, the optimising
        transformations are very likely to have introduced an error
        in the resulting program.
        %We present Core and its formal
        %basis, System~$F_C$~\cite{cite:systemfc}, in Section~\ref{sec:core}.
        % TODO: \item values in rust are linear by default while non-linear is
        % the haskell default?
%
        % \item

        Both Haskell and its intermediate language Core are \emph{lazily}
        evaluated, i.e., expressions in Haskell are only evaluated when needed,
        unlike C or Rust, which are \emph{eagerly} evaluated.
        %
        Laziness allows an optimising compiler to aggressively transform the source
        program without changing its semantics and, indeed, the Glasgow Haskell
        Compiler (GHC) heavily transforms Core by leveraging its laziness.
        %
        However, lazy evaluation interacts non-trivially with linearity.
        Intuitively, since expressions are not necessarily evaluated, an
        occurrence of a linear resource in an expression does not necessarily
        entail consuming that resource (i.e., if the expression is not evaluated,
        the resource is not used).
%
        In eagerly evaluated languages, the distinction between syntactic uses of a
        resource and the actual use of linear resources at runtime does not exist --
        an occurrence of a variable in the program always entails consuming it.
        %
        This interaction is unique to Haskell since, as far as we know, it is the
        only language featuring both laziness and linearity.
        %
        %We review lazy and eager evaluation strategies in Section~\ref{sec:background:evaluation-strategies}.

%\end{itemize}
%

@@

In spite of a linearly typed Core ultimately being the desired intermediate
language for the Glasgow Haskell Compiler,
both in its expressiveness to completely represent a Haskell with
linear types and in enabling more extensive compiler validations plus improved optimisations,
linearity is effectively ignored in Core. The reason is not evidently clear:
if we can typecheck linearity in the surface level Haskell, it should also be
possible, and natural, to do so in Core.
% why do we fail to do so in Core?

@@

\end{itemize}
%
We review background concepts fundamental to our work in
Section~\ref{sec:background}, including linear type systems, linear types in
Haskell, evaluation strategies, Core's type system, and multiple optimising
transformations applied by GHC in its compilation process.
%
We compare our contributions to related work and discuss possible
avenues for further research (highlighting so-called \emph{multiplicity
coercions}) in Section~\ref{sec:discussion}, which concludes the document.

@@

Consider the following program in a functional Haskell-like language,
where a computation that closes the given \lstinline{handle} is bound to \lstinline{x} before the
\lstinline{handle} is returned:
%
\begin{lstlisting}
f : Handle *' $\lolli$ '* Handle
f handle = let x = close handle in handle
\end{lstlisting}
%
In this example, the \lstinline{handle} appears to be closed before
returned, whereas in fact the handle will only be closed if the let bound
computation is effectively run (i.e. evaluated).
%
The example illustrates that \emph{consuming} a resource is not necessarily
synonymous with its syntactic occurrences, since depending on the evaluation strategy
of the language the computation that closes the handle might not be
executed, and if so, the \lstinline{handle} in that unused computation
is not consumed.
%
Expanding on this, consider the above example program under distinct
evaluation strategies: under call-by-value, %With \emph{eager}
% evaluation semantics,
the let
  bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
returned. It is clear that a linear type system should not accept such a
program since the linear resource \lstinline{handle} is duplicated -- it is used in a
computation that closes it, while still being made available to the caller of
the function;
%
on the other hand, with \emph{call-by-need}, the let
bound expression will only be evaluated when the binding \lstinline{x} is needed. We return
the \lstinline{handle} right away, and the let binding is forgotten as it cannot be
used outside the scope of the function, so the handle is not closed by \lstinline{f}.
Under the lens of \emph{call-by-need} evaluation, \emph{using} a resource in a
let binding only results in the resource being \emph{consumed} if the binding
itself is \emph{consumed}. We argue that a linear type system under
\emph{call-by-need} evaluation should accept the above program, unlike a linear
type system for the same program evaluated \emph{call-by-value}.

Intuitively, a computation that depends on a linear resource to produce a
result consumes that resource iff the result is effectively computed; in
contrast, a computation that depends on a linear resource, but is never run,
will not consume that resource.
%
From this observation, and exploring the connection between computation and evaluation,
% \todo{Alguma dificuldade em dizer exatamente como é que evaluation drives/is computation}
it becomes clear that \emph{linearity} and \emph{consuming resources}, in the
above example and for programs in general, should be defined in function of the
language's evaluation strategy.
%
We turn our focus to \emph{linearity} under \emph{call-by-need}, not only
because GHC Core is \emph{call-by-need}, but also because the distinction
between semantically and syntactically consuming a resource is only exposed
under \emph{non-strict} semantics.
%
% Indeed, under \emph{call-by-value}, syntactic occurrences of a linear resource
% directly correspond to semantically using that resource\footnote{With the minor exception
% of trivial aliases, which don't entail any computation even in
% \emph{call-by-value}. In theory, we could use in mutual exclusion any of the
% aliases to refer to a resource without loss of linearity} because \emph{all}
% expressions are eagerly evaluated -- if all computations are eagerly run, all
% linear resources required by computations are \emph{eagerly consumed}.

@@

\paragraph{Let bindings}
We start our discussion with non-strict (non-recursive) let bindings, i.e. let bindings whose
body is evaluated only when the binding is needed, rather than when declared.
In Core, a let binding entails the creation of a \emph{thunk} that suspends the
evaluation of the let body.
% (for background, see
% Section~\ref{sec:background:evaluation-strategies})
When the binding itself is evaluated, the \emph{thunk} is
\emph{forced} and the evaluation is carried out. The result overwrites the
\emph{thunk} -- the let binding now points to the result of the evaluation.
% A \emph{thunk} is \emph{forced} (and the suspended computation is evaluated) when
% the binding itself is evaluated.

    \begin{notyet}
\begin{lstlisting}
f1 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* b
f1 use x = let y = use x in y
\end{lstlisting}
\end{notyet}
In a linear type system, a non-strict let binding that depends on a linear
resource \lstinline{x} doesn't consume the resource as long as the binding isn't
evaluated -- the suspended computation only uses the resource if it is run. For
this reason, we can't naively tell whether \lstinline{x} is consumed just by looking at
the let binding body. In the following example, we assign a computation that
depends on the resource \lstinline{x} to a binder, which is then
returned.

% \begin{figure}
%   \begin{minipage}{.33\textwidth}
%     \begin{notyet}
% \begin{lstlisting}
% f1 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* b
% f1 use x = let y = use x in y
% \end{lstlisting}
% \end{notyet}
% \caption{one}
% \end{minipage}%
% \begin{minipage}{.33\textwidth}
%       \begin{notyet}
% \begin{lstlisting}
% f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* a
% f2 use x = let y = use x in use x
% \end{lstlisting}
% \end{notyet}
% \caption{two}
% \end{minipage}%
% \begin{minipage}{.33\textwidth}
% \begin{notyet}
% \begin{lstlisting}
% f3 :: (a *' $\lolli$ '* a) -> Bool -> a *' $\lolli$ '* a
% f3 use bool x =   let y = use x in
%    case bool of
%       True -> x
%       False -> y
% \end{lstlisting}
% \end{notyet}
% \caption{Three}
% \end{minipage}%
% \end{figure}

%
The linear resource \lstinline{x} is used exactly once, since it is used exactly once in
the body of the binding and the binding is used exactly once in the let body.
%
According to Linear Haskell's core calculus~$\lambda_{\to}^{q}$~\cite{cite:linearhaskell}, let
bound variables are annotated with a multiplicity which is multiplied (as per
the multiplicity semiring) by the multiplicity of all variables that are free
in the binder's body.
%
In short, if a let binder is linear (has
multiplicity $1$) then the linear variables free in its body are only used
once; if the let binder is unrestricted (has multiplicity $\omega$) then the
resources in its body are consumed many times, meaning no linear variables can
occur in that let binder's body.
%
Unfortunately, GHC's implementation of Linear
Haskell doesn't seem to infer multiplicities for lets yet, so while the above
program should typecheck in Linear Haskell, it is rejected by GHC.

The next example exposes the case in which the let binder is ignored in the let
body. Here, the linear resource |x| is used in |y|'s body and in the let
body, however, the resource is still used semantically linearly because |y| isn't used at all, thus |x| is consumed just once in the let body:
%
\begin{notyet}
\begin{lstlisting}
f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* a
f2 use x = let y = use x in use x
\end{lstlisting}
\end{notyet}
%
Programmers don't often write bindings that are completely unused, yet, an
optimising compiler will produce intermediate programs with unused bindings\footnote{Unused bindings are then
also dropped by the optimising compiler} 
from transformations such as inlining, which can substitute out occurrences
of the binder (e.g. |y| is inlined in the let body).

Let bindings can also go unused if they are defined before branching on case
alternatives. At runtime, depending on the branch taken, the let binding will
be evaluated only if it occurs in that branch.
Both optimising transformations (float-out), and programmers used to non-strict evaluation, can produce
programs with bindings that are selectively used in the case alternatives, for
instance:
%
\begin{notyet}
\begin{lstlisting}
f3 :: (a *' $\lolli$ '* a) -> Bool -> a *' $\lolli$ '* a
f3 use bool x = let y = use x in case bool of
      True -> x 
      False -> y
\end{lstlisting}
\end{notyet}
%
This example essentially merges \lstinline{f1} with \lstinline{f2}, using \lstinline{x} directly in one
branch and using \lstinline{y} in the other. Semantically, this program is linear because
the linear resource \lstinline{x} ends up being used exactly once in both case
alternatives, directly or indirectly.

Shifting our focus from not using a let binding to using it (more than once),
we reiterate that a let binding creates a \emph{thunk} which is only evaluated
once, and re-used subsequently. Despite the binder body only being evaluated
once, and thus its resources only used once to compute a result, we can still
only consume said result of the computation once -- perhaps surprisingly, as the
perception so far is that ``resources are consumed during computation'' and
multiple uses of the same let binder share the result that was computed only
once. Illustratively, the following program must \emph{not} typecheck:
%
\begin{noway}
\begin{lstlisting}
f4 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* (b, b)
f4 use x = let y = use x in (y, y)
\end{lstlisting}
\end{noway}
%
Intuitively, the result of the computation must also be used exactly once,
despite being effectively computed just once, because said result may still
contain (parts of) the linear resource. The trivial example is \lstinline{f4} applied to
\incode{id} -- the result of computing \incode{id x} is \incode{x}, and \incode{x} must definitely not be
shared. Indeed, if the result of the computation involving the linear resource
was, e.g., an unrestricted integer, then sharing the result would not involve consuming the
resource more than once.
% ROMES:TODO:!!!!
%\todo{this kind of hints that maybe somehow in the
%alternative scrutinizing a "trivial" atom we could make the let binding
%unrestricted after we were sure it was evaluated once, but that isn't easy
%without erasing too many things before, we kind of tried this once}
%
Concretely, the result of evaluating a let binder body using linear resources, if computed, must be
consumed exactly once, or, otherwise, we risk discarding or duplicating said resources.

Lastly, consider a program which defines two let bindings \incode{z} and \incode{y}, where
\incode{z} uses \incode{y} which in turn uses the linear resource \incode{x}:
%
\begin{noway}
\begin{lstlisting}
f5 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* ()
f5 use x = let y = use x in let z = use y in ()
\end{lstlisting}
\end{noway}
%
Even though the binding \incode{y} is used in \incode{z}, \incode{x} is still never consumed because
\incode{z} isn't evaluated in the let body, and consequently \incode{y} isn't evaluated
either -- never consuming \incode{x}. We use this example to highlight that even for
let bound variables, the syntactic occurrence of a variable isn't enough to
determine whether it is used. Instead, we ought to think of uses of \incode{y} as
implying using \incode{x}, and therefore uses of \incode{z} imply using \incode{x}, however, if
neither is used, then \incode{x} isn't used. Since \incode{x} is effectively discarded, this
example also violates linearity.

% \parawith{Summary}
The examples so far build an intuition for semantic linearity in the presence
of lazy let bindings. In essence, an unused let binding doesn't consume any
resources, and a let binding used exactly once consumes its resources exactly
once. Let binders that depend on linear resources must be used \emph{at most
once} -- let bound variables are \emph{affine} in the let body.
%
Moreover, if the let binding ($y$) isn't used in the let body, then the
resources it depends on ($\ov{x}$) must still be used -- the binding $y$ is
mutually exclusive with the resources $\ov{x}$ (for the resources to be used
linearly, either the binder occurs exactly once $y$, or the resources $\ov{x}$
do). We discuss how to encode this principle between let bindings and their dependencies using so called \emph{usage
environments}, in Section~\ref{sec:usage-environments}.
